---
title: "Binary (Dis)similarity"
execute: 
  enabled: true
julia: julia-1.10
bibliography: ../../../references.bib
---

## Outline

This page outlines how to calculate binary (dis)similarity coefficients using presence-absence matrices as outlined in @legendre2012.


## Import Required Packages

::: {.panel-tabset}

## Julia

```{julia import_packages_julia}
using VegSci
using NamedArrays
using LinearAlgebra
using CSV
using BenchmarkTools
using DataFrames
```


## R

```{r import_packages_r}
library(vegan)
library(microbenchmark)
```

:::

## Create example data

Create example data in the form of a site by species named matrix, making row (Releve2) and column (Spp7) contain all zero entries, for testing.

```{julia create_example_data}
mat = generate_test_array(rown = 20, coln = 30, meancoloccs = 10, rowprefix = "SiteA-", colprefix = "Species")
```

Create a presence-absence matrix by replacing all non-zero values with an Integer value of 1.

```{julia create_sample_data}
samp = Int.(mat .!= 0)
```

Using this prescence-absence matric we then need to create a contingency table....



## Calculate the transposed cross product

Calculate the transposed cross product. This contains the number of species shared between each of the plots and the total number of species per plot along the diagonal.

```{julia transposed_cross_product}
d = samp * transpose(samp)
```

## Extract the diagonal

Extract the diagonal of d (the transposed cross product of the sample matrix) and calculate its length.

```{julia d_diagonal}
begin
	s = Array(diag(d))
	n = length(s)
end
```

## Calculate a

Calculate a by removing the diagonal, leaving a matrix containing the shared presences only.

```{julia calculate_a}
a = d - LinearAlgebra.Diagonal(d)
```

## Calculate b
```{julia calculate_b}
b = reshape(repeat(s, n), :, n) - a
```

## Calculate c
```{julia calculate_c}
c = Array(transpose(reshape(repeat(s, n), :, n))) - a
```

## Calculate a binary (dis)similarity metric

Use a, b, and c to calculate the Jaccard similarity [@jaccard1912] between each of the sample plots, as defined in @eq-jaccard_sim.

```math
j = \frac{a}{a + b + c} = \frac{X \cap Y}{X \cup Y}
```

```{julia calculate_jaccard}
j = a ./ (a .+ b .+ c)
```

Optionally, ensure that each plot is recorded as being identical with itself.

```{julia jaccard_diag_set_one}
j_i = j + I
```

Optionally, for visual clarity view the lower (or upper) triangular portion of the similarity matrix.

```{julia jaccard_lower_triangular}
LinearAlgebra.LowerTriangular(j_i)
```

## Create a binary (dis)similarity metric function

Collate the above code into a function, which accepts a Matrix or NamedMatrix, x, and returns a Matrix or NamedMatrix containing the site x site Jaccard similarity values.

```{julia jaccard_function}
function jaccard_similarity(x::Union{Matrix, NamedMatrix})
	samp = Int.(x .!= 0)
	d = samp * transpose(samp)
	s = Array(diag(d))
	n = length(s)
	a = d - Diagonal(d)
	b = reshape(repeat(s, n), :, n) - a
	c = Array(transpose(reshape(repeat(s, n), :, n))) - a
	j = (a ./ (a .+ b .+ c)) + I
	return j
end	
```

Execute function.

```{julia jaccard_function_execute}
jaccard_similarity(mat)
```

## Store equations in object

```{julia dissim_measures_const}
similarity_indices = 
	Dict(
		"jaccard" => "(a ./ (a .+ b .+ c)) + I"
	)
```


```{julia foo}
id2 = Base.ImmutableDict("key2"=>2)
id2.key2
```

## Create a customisable (dis)similarity function

There are a number of binary (dis)similarity metrics which utlise presence-absence data, with the R package `{vegan}` supporting the 24 indices of beta diversity descibed in @koleff2003 in the function `vegan::betadiver()` and 16 (dis)similarity metrics in the function `vegan::vegdist()`. The ability to use a custom dissimilarity metric with the function `vegan::designdist()` is also provided [@oksanen2022].

The metaprogramming capabilities of Julia facilitate the implementation of a user-supplied (dis)similarity metric with ease. 
Only three additional steps are required. 
First, compose a string representing an equation. Second, create an expression by parsing this string. Last, evaluate this expression.


```{julia parse_string_jaccard}
begin
	equation = "(a ./ (a .+ b .+ c)) + I"
	expression = Meta.parse(equation)
	eval(expression)
end
```

Next create a new function with an additional parameter, eq.

```{julia custom_dissimilarity_function}
function binary_dissimilarity(x::Union{Matrix, NamedMatrix}, eq::String)
	samp = Int.(x .!= 0)
	d = samp * transpose(samp)
	s = Array(diag(d))
	n = length(s)
	a = d - Diagonal(d)
	b = reshape(repeat(s, n), :, n) - a
	c = Array(transpose(reshape(repeat(s, n), :, n))) - a
	sim = eval(Meta.parse(eq))

	f(a,b,c) = eval(Meta.parse(eq))
	sim = f(a,b,c)
	return sim
end	
```

Test this function with a different (dis)similarity metric, for example the Whittaker's beta diversity metric ($\beta_{w}$) as formulated in [@koleff2003] and defined in @eq-whittaker_beta where $S$ is equal to the total number of species recorded in both plots and $\bar{\alpha}$ is equal to the average number of species recorded in the sample plots.

```math
\beta_{w} = \frac{a + b + c}{(2a + b + c)/2} -1 = \frac{S}{\bar{\alpha}} -1
```

```{julia execute_custom_dissimilarity_function}
binary_dissimilarity(mat, "((a .+ b .+ c) ./ ((2 * a .+ b .+ c) ./ 2)) .- 1")

@eval(@formula($(Meta.parse(text))))
```

## R vs Julia Dune Data

Let's see how the R function `vegan::designdist()` compares to the Julia function `similarity()` created above.

First load the `dune` dataset bundled in `{vegan}`.

::: {.panel-tabset}

## Julia

```{julia load_dune_julia}
begin
	dune_df = CSV.read("../data/dune.csv", DataFrame, header = 1)
	dune_na = NamedArray(Array(dune_df))
	NamedArrays.setnames!(dune_na, names(dune_df), 2)
	dune_na
end
```


## R

```{r load_dune_r}
dune <- as.matrix(read.csv(file = "../data/dune.csv"))
```

:::

Next, let's benchmark both implementations using the dune data

```{julia benchmark_julia}
@benchmark similarity(x = dune_na, eq = "(a ./ (a .+ b .+ c))")
```

```{r benchmark_r}
microbenchmark::microbenchmark(vegan::designdist(x = dune, "(a)/(a+b+c)", abcd = TRUE), units = "us")
```


## Macro

```{julia macro}
macro eval_str(x)
          :(eval(Meta.parse($x)))
end
```

```{julia eval_macro}
@binarysimilarity(mat, "(a ./ (a .+ b .+ c))")
```


## References